Formalization and Automated Verification of \\ G\"{o}del's Proof of God's Existence\thanks{This work has been supported by the German Research Foundation under grant BE2501/9-1.}

### 2013-09-13 Abstract/Square/GodProof-Abstract-SquareVersion.tex
further reduction to 1 page

\usepackage{url,amsmath,amssymb,a4wide}

@@

\institute{
  Dahlem Center for Intelligent Sstems, Freie Universit\"{a}t Berlin, Germany\\
  \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
  \email{bruno@logic.at}
}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable (but debatable) axioms that they should satisfy.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof
(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography}). We have analyzed Scott's version of G\"{o}del's proof for the first-time
with an unprecedent degree of detail 
and formality with the help of theorem provers; cf.~\cite{FormalTheologyRepository}. 
The following has been done (and in this order):
\begin{itemize}
\item A detailed natural deduction proof.
%
\item A formalization of the axioms, definitions and theorems in the TPTP THF syntax \cite{J22}.
%
\item Automatic verification of the consistency of the axioms and 
definitions with Nitpick \cite{Nitpick}.
%
\item Automatic demonstration of the theorems with the provers LEO-II \cite{LEO-II} and Satallax \cite{Satallax}.

\item A step-by-step formalization using the Coq proof assistant \cite{Coq}.

\item A formalization using the Isabelle proof assistant \cite{Isabelle}, where the theorems (and some additional lemmata) have been automated with Sledgehammer \cite{Sledgehammer} and Metis \cite{Hurd03first-orderproof}.
\end{itemize}

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantifiers for individuals and properties.  Our computer-assisted formalizations rely on an
embedding of the modal logic into classical higher-order logic with
Henkin semantics \cite{J23,B9}. The formalization is thus essentially
done in classical higher-order logic where  quantified modal logic is
emulated.

In our ongoing computer-assisted study of G\"odel's proof we have
obtained the following results:
\begin{itemize}
\item The basic modal logic K is sufficient for proving T1, C and T2. 
\item Modal logic S5 is not needed for proving T3; the logic KB is
  sufficient. 
\item Without the first conjunct $\phi(x)$ in D2 the set of axioms 
  and definitions would be inconsistent.
\item For proving theorem T1, only the left to right direction of
  axiom A1 is needed. However, the backward direction of A1 is
  required for proving T2.

\end{itemize}

@@

\clearpage

@@

\bibitem{Adams}
R.M. Adams.
\newblock Introductory note to *1970.
\newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
  Letters}}. Oxford University Press, 1995.

\bibitem{AndersonGettings}
A.C. Anderson and M.~Gettings.
\newblock G\"odel ontological proof revisited.
\newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
  Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

\bibitem{B9}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Exploring properties of normal multimodal logics in simple type
  theory with {LEO-II}.
\newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
  Birthday}}, pages 386--406. College Publications.

\bibitem{J23}
C.~Benzm{\"u}ller and L.C. Paulson.
\newblock Quantified multimodal logics in simple type theory.
\newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
  7(1):7--20, 2013.

\bibitem{LEO-II}
C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke.
\newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
  logic.
\newblock In {\em Proc. of IJCAR 2008}, volume 5195 of {\em LNAI}, pages
  162--170. Springer, 2008.

\bibitem{Coq}
Y.~Bertot and P.~Casteran.
\newblock {\em {Interactive Theorem Proving and Program Development}}.
\newblock Springer, 2004.

\bibitem{Sledgehammer}
J.C. Blanchette, S.~B\"ohme, and L.C. Paulson.
\newblock Extending {Sledgehammer} with {SMT} solvers.
\newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

\bibitem{Nitpick}
J.C. Blanchette and T.~Nipkow.
\newblock Nitpick: A counterexample generator for higher-order logic based on a
  relational model finder.
\newblock In {\em Proc. of ITP 2010}, number 6172 in LNCS, pages 131--146.
  Springer, 2010.

\bibitem{Satallax}
C.E. Brown.
\newblock Satallax: An automated higher-order prover.
\newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
  Springer, 2012.

\bibitem{ContemporaryBibliography}
R.~Corazzon.
\newblock Contemporary bibligraphy on the ontological proof
  (\url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}).

\bibitem{Fitting}
M.~Fitting.
\newblock {\em {Types, Tableaux and G\"odel's God}}.
\newblock Kluver Academic Press, 2002.

@@

\bibitem{Hazen}
A.P. Hazen.
\newblock On g\"odel's ontological proof.
\newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

\bibitem{Hurd03first-orderproof}
J.~Hurd.
\newblock First-order proof tactics in higher-order logic theorem provers.
\newblock In {\em Design and Application of Strategies/Tactics in Higher Order
  Logics, NASA Tech. Rep. NASA/CP-2003-212448}, 2003.

\bibitem{Isabelle}
T.~Nipkow, L.C. Paulson, and M.~Wenzel.
\newblock {\em {Isabelle/HOL: A Proof Assistant for Higher-Order Logic}}.
\newblock Number 2283 in LNCS. Springer, 2002.

\bibitem{FormalTheologyRepository}
B.~Woltzenlogel Paleo and C.~Benzm\"uller.
\newblock Formal theology repository
  (\url{https://github.com/FormalTheology/GoedelGod}).

@@

\bibitem{J22}
G.~Sutcliffe and C.~Benzm{\"u}ller.
\newblock Automated reasoning in higher-order logic using the {TPTP THF}
  infrastructure.
\newblock {\em Journal of Formalized Reasoning}, 3(1):1--27, 2010.

>>>>

\usepackage{amsmath,amssymb,a4wide}

@@

\institute{
  Dahlem Center for Intelligent Sstems, Freie Universit\"{a}t Berlin, Germany\\
 % \email{c.benzmueller@gmail.com}
  \and 
  Theory and Logic Group, Vienna University of Technology, Austria \\
 %  \email{bruno@logic.at}
}

@@

Attempts to prove the existence (or non-existence) of God by means of
abstract ontological arguments are an old tradition in philosophy and
theology.  G\"{o}del's proof \cite{Goedel1970,GoedelNotes} is a modern culmination of
this tradition, following particularly the footsteps of Leibniz.
%
G\"{o}del defines God as a being who possesses all \emph{positive} properties.
He does not extensively discuss what positive properties are, 
but instead he states a few reasonable (but debatable) axioms that they should satisfy.
Various slightly different versions of axioms and definitions have
been considered by G\"{o}del and by several philosophers who commented
on his proof. %
%(cf. \cite{sobel2004logic,AndersonGettings,Fitting,Adams,ContemporaryBibliography})
We have analyzed Scott's version of G\"{o}del's proof \cite{ScottNotes} for the first-time
with an unprecedent degree of detail 
and formality with the help of theorem provers; cf.~\url{https://github.com/FormalTheology/GoedelGod}.

The following has been done (and in this order):\footnote{System URLs:
 TPTP---\url{http://tptp.org};
 Nitpick---\url{http://www4.in.tum.de/\~{}blanchet/nitpick.html};
 LEO-II---\url{http://leoprover.org};
 Satallax---\url{http://www.ps.uni-saarland.de/~cebrown/satallax/}; 
 Coq---\url{http://coq.inria.fr};
 Isabelle---\url{http://isabelle.in.tum.de}.
}
(i) a detailed natural deduction proof;
%
(ii) a formalization in TPTP THF syntax;
%
(iii) an automatic verification of the consistency of the axioms and 
definitions with Nitpick;
%
(iv) an automatic demonstration of the theorems with the provers LEO-II and Satallax;
%
(v) a step-by-step formalization using the Coq proof assistant;
%
(vi) a formalization using the Isabelle proof assistant, where the
theorems (and some additional lemmata) have been automated with the
Isabelle tools Sledgehammer and Metis.

G\"{o}del's proof is challenging to formalize and verify because it
requires an expressive logical language with modal operators
(\emph{possibly} and \emph{necessarily}) and with
quantifiers for individuals and properties.  Our computer-assisted formalizations rely on an
embedding of the modal logic into classical higher-order logic with
Henkin semantics. The formalization is thus essentially
done in classical higher-order logic where  quantified modal logic is
emulated.

In our ongoing computer-assisted study of G\"odel's proof our proof
tools have made
some interesting observations, including:
(a) The basic modal logic K is sufficient for proving the first three
theorems (T1, Coro and T2) as outlined in Scott's notes.
(b)  For proving the final theorem (T3), logic KB is
  sufficient.
(c) G\"odel's original  version of the proof \cite{GoedelNotes}, which omits
conjunct $\phi(x)$ in definition of \textit{essence}, seems inconsistent.
% (d) for proving theorem T1, only the left to right direction of
%   axiom A1 is needed. However, the backward direction of A1 is
%   required for proving T2.
% \end{itemize}

@@

\small

@@

% \bibitem{Adams}
% R.M. Adams.
% \newblock Introductory note to *1970.
% \newblock In {\em {Kurt G\"odel: Collected Works Vol. 3: Unpublished Essays and
%   Letters}}. Oxford University Press, 1995.

% \bibitem{AndersonGettings}
% A.C. Anderson and M.~Gettings.
% \newblock G\"odel ontological proof revisited.
% \newblock In {\em {G\"odel'96: Logical Foundations of Mathematics, Computer
%   Science, and Physics: Lecture Notes in Logic 6}}. {Springer}, 1996.

% \bibitem{B9}
% C.~Benzm{\"u}ller and L.C. Paulson.
% \newblock Exploring properties of normal multimodal logics in simple type
%   theory with {LEO-II}.
% \newblock In {\em {Festschrift in Honor of {Peter B. Andrews} on His 70th
%   Birthday}}, pages 386--406. College Publications.

% \bibitem{J23}
% C.~Benzm{\"u}ller and L.C. Paulson.
% \newblock Quantified multimodal logics in simple type theory.
% \newblock {\em Logica Universalis (Special Issue on Multimodal Logics)},
%   7(1):7--20, 2013.

% \bibitem{LEO-II}
% C.~Benzm{\"u}ller, F.~Theiss, L.~Paulson, and A.~Fietzke.
% \newblock {LEO-II} - a cooperative automatic theorem prover for higher-order
%   logic.
% \newblock In {\em Proc. of IJCAR 2008}, volume 5195 of {\em LNAI}, pages
%   162--170. Springer, 2008.

% \bibitem{Coq}
% Y.~Bertot and P.~Casteran.
% \newblock {\em {Interactive Theorem Proving and Program Development}}.
% \newblock Springer, 2004.

% \bibitem{Sledgehammer}
% J.C. Blanchette, S.~B\"ohme, and L.C. Paulson.
% \newblock Extending {Sledgehammer} with {SMT} solvers.
% \newblock {\em Journal of Automated Reasoning}, 51(1):109--128, 2013.

% \bibitem{Nitpick}
% J.C. Blanchette and T.~Nipkow.
% \newblock Nitpick: A counterexample generator for higher-order logic based on a
%   relational model finder.
% \newblock In {\em Proc. of ITP 2010}, number 6172 in LNCS, pages 131--146.
%   Springer, 2010.

% \bibitem{Satallax}
% C.E. Brown.
% \newblock Satallax: An automated higher-order prover.
% \newblock In {\em Proc. of IJCAR 2012}, number 7364 in LNAI, pages 111 -- 117.
%   Springer, 2012.

% \bibitem{ContemporaryBibliography}
% R.~Corazzon.
% \newblock Contemporary bibligraphy on the ontological proof
%   (\url{http://www.ontology.co/biblio/ontological-proof-contemporary-biblio.htm}).

% \bibitem{Fitting}
% M.~Fitting.
% \newblock {\em {Types, Tableaux and G\"odel's God}}.
% \newblock Kluver Academic Press, 2002.

@@

% \bibitem{Hazen}
% A.P. Hazen.
% \newblock On g\"odel's ontological proof.
% \newblock {\em Australasian Journal of Philosophy}, 76:361--377, 1998.

% \bibitem{Hurd03first-orderproof}
% J.~Hurd.
% \newblock First-order proof tactics in higher-order logic theorem provers.
% \newblock In {\em Design and Application of Strategies/Tactics in Higher Order
%   Logics, NASA Tech. Rep. NASA/CP-2003-212448}, 2003.

% \bibitem{Isabelle}
% T.~Nipkow, L.C. Paulson, and M.~Wenzel.
% \newblock {\em {Isabelle/HOL: A Proof Assistant for Higher-Order Logic}}.
% \newblock Number 2283 in LNCS. Springer, 2002.

% \bibitem{FormalTheologyRepository}
% B.~Woltzenlogel Paleo and C.~Benzm\"uller.
% \newblock Formal theology repository
%   (\url{https://github.com/FormalTheology/GoedelGod}).

@@

% \bibitem{J22}
% G.~Sutcliffe and C.~Benzm{\"u}ller.
% \newblock Automated reasoning in higher-order logic using the {TPTP THF}
%   infrastructure.
% \newblock {\em Journal of Formalized Reasoning}, 3(1):1--27, 2010.
